### 2021-07-07 rad-icfp-2021/purerad.tex
.

In Figure~\ref{fig:direct_diff_macro} we present our direct transformation from the source language to the target language for pure reverse mode differentiation.
We explain in the next Section (\ref{sub:Partial evaluation and optimization}) how to optimize it further and compute the gradient.
Admittedly, this transformation may be hard to read, and it is not straightforward to show its correctness directly. 
In Section~\ref{sec:unf} we decompose this transformation into three simpler steps via a novel intermediate representation (UNF).

@@

\begin{notation}
We now introduce several notations which are useful when defining the transformation for reverse-mode in Figure~\ref{tbl:notation:one}.
Ground types are defined inductively by 
$$G::= \RR \mid G\times \ldots \times G \mid A[G]$$

@@

\begin{tabular}{l c l}
    $0_\RR$  & $\defeq$ & $\underline{0}$ \\
    $0_{G_1\times \ldots \times G_n}$ & $\defeq$ &  $< O_{G_1},\ldots, 0_{G_n} >$ \\
    $0_{\Array{\reals}{n}}$& $\defeq$ & !ZerosLike(n)! \\
    $a\widehat{+}_\RR b$ & $\defeq$ & $ a+b$ \\
    $(a_1,\ldots,a_n)\widehat{+}_{G_1\times\ldots\times G_n}(b_1,\ldots,b_n)$ & $\defeq$ & $(a_1\widehat{+}_{G_1}b_1,\ldots,a_n\widehat{+}_{G_n}b_n)$ \\
    $A\widehat{+}_{A[\RR]}B $ & $\defeq$ & !map2! + $A$ $B$ 
\end{tabular}

@@

We have the following typing lemma for $\directD{\rho}{\Gamma}{Y}$.
\begin{lemma}[Typing $\directD{\rho}{\Gamma}{Y}$]
    If $\Gamma \vdash$ !e: A!, then $\Gamma$!,Y: !$\Gamma\to \rho \vdash \directD{\rho}{\Gamma}{Y}$!(e): !$\directD{\rho}{\Gamma}{Y}$!(A)!.
\end{lemma}

@@

Given a term $\Gamma\vdash e : \reals$, we can compute its gradient $\grad_\Gamma e$ from a particular instance of 
$\directD{\rho}{\Gamma}{Y}(e)$. First, $\rho, Y$ specifies if we want to compute the whole gradient regarding the variables from $\Gamma$ or a subset of it.
For a subset $\rho\subset \Gamma$, one chooses $Y$ to be the projection function sending a variable 
$x_i:G$ of $\Gamma$ to $x_i$ if it belongs to $\rho$ and to $0_G$ otherwise.
In particular, we take $Y=Id_\Gamma$ to compute the whole gradient.
Next, the gradient will be given by the second part of the pair $\directD{\rho}{\Gamma}{Y}(e)$, 
and we need to initialize the tangent variables. All of them are set to $0$ except the one corresponding to the output value of !e!, 
which we initialize at $1$ to run the backpropagation. 
All in all, we compute the gradient via $\pi_2\directD{\rho}{\Gamma}{Id_\Gamma}(e)(0_\Gamma,1)$.

Following the insight from Section~\ref{subsec:insights}, 
we don't want to keep all the lambda abstractions as this is costly. 
The transformation is designed in such a way that all the lambda abstractions are given arguments,  
and we can use partial evaluation to do beta-reduce all these lambda abstractions.

By inspection in Figure~\ref{fig:direct_diff_macro}, 
we see that there is a linear usage of  each continuation $Y$ 
and that it is always applied to almost the identity. 
More precisely, we have applications of the form !fun (x1,$\ldots$,xn) -> Y(e1,$\ldots$,en)! 
where the !ei! are !xi! except for at most $k$ (independant of $n$) terms.
In fact we have $k=2$, except for the !map2! case which requires a more detailed analysis (see Section~\ref{sec:complexity}).

Using inlining as given in Figure~\ref{fig:optim}, this allows us to rewrite

@@

Using the invariant above, most of the !e!$_{i}$ are variables. 
Without loss of generality, assume the onlt terms !e!$_{i}$ which are not variables are !e!$_{n-1}$ and !e!$_{n}$.
we can use forward subsitution on the other !e!$_{i}$ and we obtain

@@

>>>>

In Figure~\ref{fig:direct_diff_macro} we present our direct transformation from the source language to the target language for pure reverse mode differentiation.
Given a term $\Gamma\vdash e : \reals$, we can compute its gradient $\grad_\Gamma e$ from a particular instance of 
$\directD{\rho}{\Gamma}{Y}(e)$. First, $\rho, Y$ specifies if we want to compute the whole gradient regarding the variables from $\Gamma$ or a subset of it.
For a subset $\rho\subset \Gamma$, one chooses $Y$ to be the projection function sending a variable 
$x_i:G$ of $\Gamma$ to $x_i$ if it belongs to $\rho$ and to $0_G$ otherwise.
In particular, we take $Y=Id_\Gamma$ to compute the whole gradient.
Next, the gradient will be given by the second part of the pair $\directD{\rho}{\Gamma}{Y}(e)$, 
and we need to initialize the tangent variables. All of them are set to $0$ except the one corresponding to the output value of !e!, 
which we initialize at $1$ to run the backpropagation. 
All in all, we compute the gradient via $\pi_2\directD{\rho}{\Gamma}{Id_\Gamma}(e)(0_\Gamma,1)$.

@@

\begin{notation}
We now introduce several notations which are useful when defining the transformation for reverse-mode in Figure~\ref{tbl:notation:one}.
Ground types are defined inductively by 
$$G::= \RR \mid G\times \ldots \times G \mid \Array{\RR}{n}$$

@@

\begin{tabular}{l c l}
    $0_\RR$  & $\defeq$ & $\underline{0}$ \\
    $0_{G_1\times \ldots \times G_n}$ & $\defeq$ &  $< O_{G_1},\ldots, 0_{G_n} >$ \\
    $0_{\Array{\RR}{n}}$& $\defeq$ & !ZerosLike(n)! \\
    $a\widehat{+}_\RR b$ & $\defeq$ & $ a+b$ \\
    $(a_1,\ldots,a_n)\widehat{+}_{G_1\times\ldots\times G_n}(b_1,\ldots,b_n)$ & $\defeq$ & $(a_1\widehat{+}_{G_1}b_1,\ldots,a_n\widehat{+}_{G_n}b_n)$ \\
    $A\widehat{+}_{\Array{\RR}{n}}B $ & $\defeq$ & !map2! + $A$ $B$ 
\end{tabular}

@@

We have the following typing lemma for $\directD{\rho}{\Gamma}{Y}$.
\begin{lemma}[Typing $\directD{\rho}{\Gamma}{Y}$]
    If $\Gamma \vdash$ !e: A!, then $\Gamma$!,Y:!$\Gamma\to \rho \vdash \directD{\rho}{\Gamma}{Y}$!(e): !$\directD{\rho}{\Gamma}{Y}$!(A)!.
\end{lemma}
\begin{proof}
By induction on derivation of $\Gamma\vdash$!e:A!.
\end{proof}

@@

Following the insight from Section~\ref{subsec:insights}, 
we do not want to keep all the lambda abstractions as this is costly. 
The transformation is designed in such a way that all the lambda abstractions are given arguments,  
and we can use partial evaluation to do beta-reduce all these lambda abstractions.

By inspection in Figure~\ref{fig:direct_diff_macro}, 
we see that there is a linear usage of  each continuation $Y$ 
and that it is always applied to almost the identity. 
More precisely, we have applications of the form !fun (x1,$\ldots$,xn) -> Y(e1,$\ldots$,en)! 
where the !ei! are !xi! except for at most $k$ (independent of $n$) terms.
In fact we have $k=2$, except for the !map2! case which requires a more detailed analysis (see Section~\ref{sec:complexity}).

Using inlining as given in Figure~\ref{fig:optim} allows us to rewrite

@@

Using the invariant above, most of the !e!$_{i}$ are variables. 
Without loss of generality, assume the only terms !e!$_{i}$ which are not variables are !e!$_{n-1}$ and !e!$_{n}$.
we can use forward substitution on the other !e!$_{i}$ and we obtain

@@

Admittedly, the transformation presented in this section may be hard to read, and it is not straightforward to show its correctness directly. 
Next, we decompose this transformation into three simpler steps via a novel intermediate representation (UNF).
